Tropical Andes Taxonomic Diversity of Plants and Frugivores

### 2024-05-10 code/L1/L1_P2_TropicalAndes_taxcleaning.Rmd

# Disparties
all_species_no.opinion <- subset(all_species_hamonized_results, Taxonomic_status == "No opinion")
nrow(all_species_no.opinion)
all_species_illegitimate <- subset(all_species_hamonized_results, Taxonomic_status == "Illegitimate")
nrow(all_species_illegitimate)
all_species_invalid <- subset(all_species_hamonized_results, Taxonomic_status == "Invalid")
nrow(all_species_invalid)
all_species_other <- subset(all_species_hamonized_results, Taxonomic_status == "")
nrow(all_species_other)
all_species_NA <- subset(all_species_hamonized_results, is.na(all_species_hamonized_results$Taxonomic_status))
nrow(all_species_NA)
```
# Create species list of names with disparities
```{r}
all_species_disparities <- all_species_hamonized_results %>%
                            filter(Taxonomic_status %in% c("No opinion", "Illegitimate", "Invalid", "", NA)) %>%
                            select(Name_submitted)
nrow(all_species_disparities)
```

>>>>

# Disparties
all_species_no.opinion <- subset(all_species_hamonized_results, Taxonomic_status == "No opinion")
nrow(all_species_no.opinion)
all_species_illegitimate <- subset(all_species_hamonized_results, Taxonomic_status == "Illegitimate")
nrow(all_species_illegitimate)
all_species_invalid <- subset(all_species_hamonized_results, Taxonomic_status == "Invalid")
nrow(all_species_invalid)
all_species_other <- subset(all_species_hamonized_results, Taxonomic_status == "")
nrow(all_species_other)
all_species_NA <- subset(all_species_hamonized_results, is.na(all_species_hamonized_results$Taxonomic_status))
nrow(all_species_NA)
```
# Create species list of names with disparities
```{r}
all_species_disparities <- all_species_hamonized_results %>%
                            filter(Taxonomic_status %in% c("No opinion", "Illegitimate", "Invalid", "", NA)) %>%
                            dplyr::select(Name_submitted)
nrow(all_species_disparities)
```

### 2024-05-10 code/L1/L1_P3_TropicalAndes_plant_traits.Rmd

>>>>

# Combine all traits
```{r}
# add unit column to categorical traits
cat_traits <- rbind(averagetraits_dispersal, averagetraits_fruitType, averagetraits_growthForm)
cat_traits$Unit <- ""
clean_traits <- rbind(averagetraits_fruitLength, averagetraits_fruitMass, averagetraits_plantHeight, averagetraits_plantLifespan, averagetraits_seedLength, averagetraits_seedMass, averagetraits_seedWidth, cat_traits)
nrow(clean_traits)
```

```{r}
clean_traits %>%
  count(TraitName)
```

# Summary
```{r}
print("Number of records: ")
nrow(clean_traits)
print("Number of species: ")
length(unique(clean_traits$Accepted_species))
```

# Get one trait value per species
```{r}
#remove unit column from dataframe
clean_traits <- clean_traits[, !colnames(clean_traits) %in% "Unit", drop = FALSE]
```

```{r}
# separate numeric and non-numeric traits
numeric_traits <- clean_traits[clean_traits$TraitName %in% c("Fruit Length", "Fruit Mass", "Plant Height", "Plant Lifespan", "Seed Length", "Seed Width", "Seed Mass"), ]
cat_traits <- clean_traits[clean_traits$TraitName %in% c("Dispersal Syndrome", "Fruit Type", "Growth Form"), ]

numeric_traits$TraitValue <- as.numeric(numeric_traits$TraitValue)
```

```{r}
# numeric traits
## Group by species and TraitName, and calculate the mean TraitValue for each combination
numeric_traits_combined <- numeric_traits %>%
  group_by(species, TraitName) %>%
  summarise(TraitValue = mean(TraitValue, na.rm = TRUE)) %>%
  ungroup()

# non-numeric traits
cat_traits_combined <- cat_traits %>%
  group_by(species) %>%
  summarise_all(.funs = list(mode_trait = function(x) {
    mode_value <- names(sort(table(x), decreasing = TRUE))[1]  # Calculate mode
    return(mode_value)
  }))
```

```{r}
#rename columns to original values
colnames(cat_traits_combined)[which(colnames(cat_traits_combined) == "TraitValue_mode_trait")] <- "TraitValue"
colnames(cat_traits_combined)[which(colnames(cat_traits_combined) == "TraitName_mode_trait")] <- "TraitName"

# Remove rows with "?" or numbers in the TraitValue column
cleaned_cat_traits <- cat_traits_combined %>%
  filter(!grepl("\\?", TraitValue) & !grepl("\\d", TraitValue))
```

```{r}
#combine numeric and non-numeric traits
combined_clean_traits <- rbind(cat_traits_combined, numeric_traits_combined)
```

```{r}
combined_clean_traits %>%
  count(TraitName)
```

# Remove traits with less than 10 records
```{r}
combined_clean_traits <- combined_clean_traits[!(combined_clean_traits$TraitName %in% c("Fruit Mass", "Seed Mass")), ]
```

# Transform plant dataframe from long to wide
```{r}
wide_traits <- spread(combined_clean_traits, key = TraitName, value = TraitValue)

#remove X column from dataframe
wide_plant_traits <- wide_traits[, !colnames(wide_traits) %in% "X", drop = FALSE]
```

```{r}
# Assign data types
wide_plant_traits$`Fruit Length` <- as.numeric(wide_plant_traits$`Fruit Length`)
wide_plant_traits$`Plant Height` <- as.numeric(wide_plant_traits$`Plant Height`)
wide_plant_traits$`Dispersal Syndrome` <- as.character(wide_plant_traits$`Dispersal Syndrome`)
wide_plant_traits$`Growth Forme` <- as.character(wide_plant_traits$`Growth Form`)
wide_plant_traits$`Fruit Type` <- as.character(wide_plant_traits$`Fruit Type`)
```

# Summary
```{r}
print("Number of species: ")
length(unique(wide_plant_traits$species))
```

# Write data to csv
```{r}
write.csv(wide_plant_traits, file.path(output_path,"TropicalAndes_all_plant_traits_harmonized.csv"))
write.csv(clean_traits, file.path(output_path,"TropicalAndes_all_plant_traits_harmonized_with_units.csv"))
```

### 2024-05-10 code/L2/L2_1_TropicalAndes_taxdiv.Rmd

  # Plot Empty Grid
  gridPlot <- ggplot() +
    geom_sf(data = Americas, fill = "white") +
    geom_sf(data = TropicalAndes_IUCNHabitat_Forest) +
    geom_sf(data = TAGrid) +
    labs(title = paste("Tropical Andes Grid [", resolution_meters/1000, " km]", sep=""), x = "Latitude", y = "Longitude") +
    coord_sf(xlim = c(-85, -54), ylim = c(-24, 14), expand = FALSE, crs = 4326) +
    annotation_scale(location = "bl", width_hint = 0.5) +
    annotation_north_arrow(location = "bl", which_north = "true", pad_x = unit(0.3, "in"), 
                           pad_y = unit(0.3, "in"), style = north_arrow_fancy_orienteering) +
    theme(panel.background = element_rect(fill = "lightblue"))

  print(gridPlot)

@@

  # Plot richness for plants
  plantgridRichTA <- ggplot(plant_richness_grid) +
    geom_sf(data = Americas, fill = "white") +
    geom_sf(data = TApoly, fill = "grey", size = 0.1) +
    geom_sf(aes(fill = num_species), color = NA) +
    scale_fill_scico(palette = "romaO", direction = 1) +
    labs(title = paste("Tropical Andes Plant Richness [", resolution_meters/1000, " km]", sep=""), x = "Longitude", y = "Latitude",
         fill = "Number of species") +
    coord_sf(xlim = c(-85, -54), ylim = c(-24, 14), expand = FALSE, crs = 4326) +
    annotation_scale(location = "bl", width_hint = 0.5) +
    annotation_north_arrow(location = "bl", which_north = "true", 
                           pad_x = unit(0.3, "in"), pad_y = unit(0.3, "in"), style = north_arrow_fancy_orienteering) +
    theme(panel.background = element_rect(fill = "lightblue"))

  print(plantgridRichTA)

  # Plot richness for frugivores
  frugivoregridRichTA <- ggplot(frugivore_richness_grid) +
    geom_sf(data = Americas, fill = "white") +
    geom_sf(data = TApoly, fill = "grey", size = 0.1) +
    geom_sf(aes(fill = num_species), color = NA) +
    scale_fill_scico(palette = "romaO", direction = 1) +
    labs(title = paste("Tropical Andes Frugivore Richness [", resolution_meters/1000, " km]", sep=""), x = "Longitude", y = "Latitude",
         fill = "Number of species") +
    coord_sf(xlim = c(-85, -54), ylim = c(-24, 14), expand = FALSE, crs = 4326) +
    annotation_scale(location = "bl", width_hint = 0.5) +
    annotation_north_arrow(location = "bl", which_north = "true", 
                           pad_x = unit(0.3, "in"), pad_y = unit(0.3, "in"), style = north_arrow_fancy_orienteering) +
    theme(panel.background = element_rect(fill = "lightblue"))

  print(frugivoregridRichTA)

  # Extract cell values
  plant_cellRichness <- plantgridRichTA[["data"]]
  frugivore_cellRichness <- frugivoregridRichTA[["data"]]

  ### Plant vs frugivore richness plot
  compare_richness_summary <- summary(lm(frugivore_cellRichness$num_species ~ plant_cellRichness$num_species))
  print(compare_richness_summary)

  # save r-squared value
  compare_richness_r2 <- compare_richness_summary$r.squared

  plant_frug_richness_compare <- plot(plant_cellRichness$num_species, frugivore_cellRichness$num_species,
       main = paste("Plant Richness vs Frugivore Richness [", resolution_meters/1000, " km]", sep = ""),
       xlab = "Plant Richness by cell", ylab = "Frugivore Richness by cell")
  abline(lm(frugivore_cellRichness$num_species ~ plant_cellRichness$num_species))

  print(plant_frug_richness_compare)

  plant_richness_hist <- hist(plant_cellRichness$num_species,
       main = paste("Histogram of Plant Richness [", resolution_meters/1000, " km] by cell", sep = ""),
       xlab = "Plant Richness by cell")
  print(plant_richness_hist)
  frugivore_richness_hist <- hist(frugivore_cellRichness$num_species,
       main = paste("Histogram of Frugivore Richness [", resolution_meters/1000, " km] by cell", sep = ""),
       xlab = "Frugivore Richness by cell")
  print(frugivore_richness_hist)
}

@@

## 1 km resolution
```{r}
richness_1km <- create_rich_plots(1000)
# save plots
plantgridRichTA_1km <- plantgridRichTA
frugivoregridRichTA_1km <- frugivoregridRichTA
# save richness data
plant_cellRichness_1km <- plant_cellRichness
frugivore_cellRichness_1km <- frugivore_cellRichness

compare_richness_r2_1km <- compare_richness_r2
```

>>>>

@@

  # Helper function to generate plots
  generate_plot <- function(data, title) {
    ggplot(data) +
      geom_sf(data = Americas, fill = "white") +
      geom_sf(data = TApoly, fill = "grey", size = 0.1) +
      geom_sf(aes(fill = num_species), color = NA) +
      scale_fill_scico(palette = "romaO", direction = 1) +
      labs(title = title, x = "Longitude", y = "Latitude", fill = "Number of species") +
      coord_sf(xlim = c(-85, -54), ylim = c(-24, 14), expand = FALSE, crs = 4326) +
      annotation_scale(location = "bl", width_hint = 0.5) +
      annotation_north_arrow(location = "bl", which_north = "true", 
                             pad_x = unit(0.3, "in"), pad_y = unit(0.3, "in"), style = north_arrow_fancy_orienteering) +
      theme(panel.background = element_rect(fill = "lightblue"))
  }

  # Generate plots
  plant_plot <- generate_plot(plant_richness_grid, paste("Tropical Andes Plant Richness [", resolution_meters/1000, " km]", sep=""))
  frugivore_plot <- generate_plot(frugivore_richness_grid, paste("Tropical Andes Frugivore Richness [", resolution_meters/1000, " km]", sep=""))

  # Compare plant and frugivore richness
  compare_richness_summary <- summary(
    lm(frugivore_richness_grid$num_species ~ plant_richness_grid$num_species))

  comparison_data <- data.frame(Plant_Richness = plant_richness_grid$num_species,
                               Frugivore_Richness = frugivore_richness_grid$num_species)

  # Create scatter plot
  plant_frug_richness_compare <- ggplot(comparison_data,
                                        aes(x = Plant_Richness, y = Frugivore_Richness)) +
    geom_point() +
    labs(title = paste("Plant Richness vs Frugivore Richness [", resolution_meters/1000, " km]",
                       sep = ""), x = "Plant Richness by cell", y = "Frugivore Richness by cell") +
    geom_smooth(method = "lm", se = FALSE, color = "red")

  # Create histograms
  plant_richness_hist <- hist(plant_richness_grid$num_species, 
    main = paste("Histogram of Plant Richness [", resolution_meters/1000, " km] by cell", sep = ""),
                              xlab = "Plant Richness by cell")
  frugivore_richness_hist <- hist(frugivore_richness_grid$num_species,
    main = paste("Histogram of Frugivore Richness [", resolution_meters/1000, " km] by cell", sep = ""),
                                  xlab = "Frugivore Richness by cell")

  # Return a list of plots
  list(plant_plot = plant_plot,
       frugivoregridRichTA = frugivore_plot,
       plant_cellRichness = plant_richness_grid,
       frugivore_cellRichness = frugivore_richness_grid, 
       compare_richness_summary = compare_richness_summary,
       plant_frug_richness_compare = plant_frug_richness_compare,
       plant_richness_hist = plant_richness_hist,
       frugivore_richness_hist = frugivore_richness_hist)
}

@@

## 1 km resolution
```{r}
richness_1km <- create_rich_plots(1000)
# save plots
plantgridRichTA_1km <- plant_plot
frugivoregridRichTA_1km <- frugivoregridRichTA
# save richness data
plant_cellRichness_1km <- plant_cellRichness
frugivore_cellRichness_1km <- frugivore_cellRichness

compare_richness_r2_1km <- compare_richness_summary$r.squared
```
